The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2017-03-17 manuscript/manuscript.Rnw
line break on -
move where acronyms are defined
add derek's suggested changes; more completely discuss ERV capture

%\documentclass[english,12pt]{article}
%\documentclass[5p]{elsarticle_dla}
\documentclass{elsarticle_dla}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSerif}
\setmathsf{FreeSans}
\setmathtt{FreeSerif}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
%\usepackage[letterpaper]{geometry}
%\usepackage{url}
\usepackage{siunitx}
\usepackage{graphicx}
%\usepackage{rotating}
%\usepackage{wrapfig}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{multirow}
%\usepackage{setspace}
\usepackage[macros=false,single=true,single-form=long]{acro}
\usepackage{array}
\usepackage{longtable}
\usepackage[export]{adjustbox}[2011/08/13]
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue,breaklinks=true]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
% \renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
\newcommand{\PP}[1]{\textcolor{orange}{\fxnote{PP: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
% \usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
\def\UrlBreaks{\do\.\do\@\do\\\do\/\do\!\do\_\do\|\do\;\do\>\do\]%
\do\)\do\,\do\?\do\&\do\'\do+\do\=\do\#%
\do\/%
}%
\input{common_latex}
\tolerance=5000
\emergencystretch=0.1pt

@@

These observations lead to two related questions: 1) what are the
genes which are responsible for the \textbf{core functionality of the
  placenta} and 2) which genes are associated with \textbf{changes in
  placenta morphology}?

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
For example, each superorder (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha, \cref{fig:species_tree})
has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
related to the thymus-specific serine protease \ac{PRSS16} in
\elephant, and the expansion of the growth hormone and
chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

@@

Fetal tissue was collected from the placenta of nine mammalian species
(\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog
[domestic dog], \cow [domestic cow], \elephant [African savanna
elephant], \opossum [gray short-tailed opossum], \galili [blind mole
rat from Upper Galilee], \carmeli [blind mole rat from northern
Israel] \cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}; n=1
for all collected species). Information on specimen ID, number of
specimens, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother and fetus from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal villous portion of
the placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). The placenta membranes were separated
from the fetal villous portion by dissection, and the
resulting fetal villous portion was washed twice in PBS. All tissues
were collected and stored in RNAlater (Life Technologies, Carlsbad,
CA) until RNA isolation could be performed. Tissues were homogenized
in Trizol, and total RNA was isolated following the manufacturer's
instructions. RNA was further purified using RNeasy Midi Kit (Qiagen,
Valencia, CA), in conjunction with the RNase-Free DNase Set (Qiagen,
Valencia, CA), following the manufacturer’s protocol to remove any
contaminating DNA from the sample. Total RNA was then evaluated on a
Nanodrop 1000 (Thermo Scientific, Wilmington, DE) and an Agilent
Bioanalyzer 2100 (Santa Clara, CA) to confirm concentration and
quality of the isolation. RNA quality was assessed using A260/A280 and
A260/A230 absorbance values.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human (\smonkey) or \mouse (\galili and \carmeli) transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}

@@

Orthologous pairs between \human and all other species were identified
using the \ac{OMA} database
\cite{Altenhoff.ea2015:OMAorthologydatabasein}. In cases where a gene
did not have a human ortholog pair identified by \ac{OMA}, Ensembl v.80
one2one orthologs obtained from Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration} %
were used to select the appropriate orthologous pair. A total of
\Sexpr{combined.fpkm[,list(n=length(gene_id),
  max.fpkm=max(mean_fpkm)),by=human_name][n==14 & max.fpkm >
  10,length(human_name)]} ortholog pair sets in all 14 species were
expressed (\ac{FPKM} $\ge 10$) in at least one species. We also
obtained orthology from the perspective of each species to all other
species using biomaRt package and Ensembl v.80 database to determine
the percentage of highly expressed genes which had an orthology of
one-to-one, one-to-many, many-to-many, or novel/unknown. For future
analyses by other investigators, we have also provided the human gene
which most closely aligns to each expressed annotated
protein-generating sequence of all species analyzed using Diamond
\cite{Buchfink.ea2015:Fastandsensitiveprotein}.

@@

All of the code used to produce every analysis presented in this paper
and the manuscript itself is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM} values are in
all\_species\_per\_sample\_fpkm.txt.xz.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \ac{ANXA2}, \ac{ANXA1},
\ac{S100A11}, \ac{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\ac{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads}.

@@

While Syncytin-1 and -2 (\ac{ERVW-1} and \ac{ERVFRD-1}, respectively) are implicated
in the formation of the syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and genes which are similar to 
\ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most of the
placentas analyzed here, they are not expressed in \bonobo, \sheep,
and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
and syncytiotrophoblast must be continuously replenished
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast should be highly expressed in the core
placenta. \ac{ANXA5} is known to be expressed at multiple stages of
placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. While there may be other genes which
are responsible for the positioning of cytotrophoblast cells in close
proximity to start the fusion process, \ac{ANXA5} is likely to
be essential for the completion of fusion and the maintenance of
membrane repair during the massive flux of nutrients and waste in the
placenta \cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing,Ueki.ea2012:Lossofmaternalannexin}.

@@

This study includes only term placentas, and our conclusions are
limited to the set of genes which continue to be expressed at term. We
suspect that many genes which are responsible for establishing shape,
intimacy, and interdigitation are expressed and function at earlier
time points where placental morphology is being established.
Furthermore, while these samples have been dissected to only include
the fetal interface, they likely include substantial cell and tissue
heterogeneity. We are therfore not yet able to assign specific genes
in our core placental transcriptome to specific cell types and
tissues. We believe that samples of placentas from earlier in
gestation and from multiple locations within the placenta and at
multiple will address these limitations.

@@

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance, such as the annexins, and genes responsible
for cell-cell and cell-matrix interactions, such as \ac{LGALS1}. These
genes were almost certainly present in the placenta of the most recent
common mammalian ancestor of the species included in this study. We hypothesize that
modifications to the expression or function of genes in the core placenta transcriptome
will compromise the
development and organization of the placenta.

>>>>

%\documentclass[english,12pt]{article}
%\documentclass[5p]{elsarticle_dla}
\documentclass{elsarticle_dla}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSerif}
\setmathsf{FreeSans}
\setmathtt{FreeSerif}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
%\usepackage[letterpaper]{geometry}
%\usepackage{url}
\usepackage{siunitx}
\usepackage{graphicx}
%\usepackage{rotating}
%\usepackage{wrapfig}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{multirow}
%\usepackage{setspace}
\usepackage[macros=false,single=true,single-form=long]{acro}
\usepackage{array}
\usepackage{longtable}
\usepackage[export]{adjustbox}[2011/08/13]
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue,breaklinks=true]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
% \renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
\newcommand{\PP}[1]{\textcolor{orange}{\fxnote{PP: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
% \usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
\def\UrlBreaks{\do\.\do\@\do\\\do\/\do\!\do\_\do\|\do\;\do\>\do\]%
\do\)\do\,\do\?\do\&\do\'\do+\do\=\do\#%
\do\/\do\-%
}%
\input{common_latex}
\tolerance=5000
\emergencystretch=0.1pt

@@

These observations lead to two related questions: 1) what are the
genes which are responsible for the \textbf{core functions of the
  placenta} and 2) which genes are associated with \textbf{changes in
  placenta morphology}?

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
For example, each superorder (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha, \cref{fig:species_tree}) has evolved
expansive placenta-specific gene families with expression patterns
that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, Pregnancy-Specific Glycoproteins in \human and \mouse,
Pregnancy-Associated Glycoproteins in \cow, genes related to the
thymus-specific serine protease \ac{PRSS16} in \elephant, and the
expansion of the growth hormone and chorionic gonadotropin beta gene
families in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

@@

Fetal tissue was collected from the placenta of nine mammalian species
(\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog
[domestic dog], \cow [domestic cow], \elephant [African savanna
elephant], \opossum [gray short-tailed opossum], \galili [blind mole
rat from Upper Galilee], \carmeli [blind mole rat from northern
Israel] \cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}; n=1
for all collected species). Information on specimen ID, number of
specimens, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother and fetus from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of
the placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). The placenta membranes were separated
from the fetal portion by dissection, and the
resulting fetal portion was washed twice in PBS. All tissues
were collected and stored in RNAlater (Life Technologies, Carlsbad,
CA) until RNA isolation could be performed. Tissues were homogenized
in Trizol, and total RNA was isolated following the manufacturer's
instructions. RNA was further purified using RNeasy Midi Kit (Qiagen,
Valencia, CA), in conjunction with the RNase-Free DNase Set (Qiagen,
Valencia, CA), following the manufacturer’s protocol to remove any
contaminating DNA from the sample. Total RNA was then evaluated on a
Nanodrop 1000 (Thermo Scientific, Wilmington, DE) and an Agilent
Bioanalyzer 2100 (Santa Clara, CA) to confirm concentration and
quality of the isolation. RNA quality was assessed using A260/A280 and
A260/A230 absorbance values.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human (\smonkey) or \mouse (\galili and \carmeli) transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length,tab:ref_genome_aligned_species})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}

@@

Orthologous pairs between \human and all other species were identified
using the \ac{OMA} database
\cite{Altenhoff.ea2015:OMAorthologydatabasein}. In cases where a gene
did not have a human ortholog pair identified by \ac{OMA}, Ensembl v.80
one2one orthologs obtained from Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration} %
were used to select the appropriate orthologous pair. A total of
\Sexpr{combined.fpkm[,list(n=length(gene_id),
  max.fpkm=max(mean_fpkm)),by=human_name][n==14 & max.fpkm >
  10,length(human_name)]} ortholog pair sets in all 14 species were
expressed (\ac{FPKM} $\ge 10$) in at least one species. We also
obtained orthology from the perspective of each species to all other
species using the biomaRt package and the Ensembl v.80 database to determine
the percentage of highly expressed genes that had an orthology of
one-to-one, one-to-many, many-to-many, or novel/unknown. For future
analyses by other investigators, we have also provided the human gene
which most closely aligns to each expressed annotated
protein-generating sequence of all species analyzed using Diamond
\cite{Buchfink.ea2015:Fastandsensitiveprotein}.

@@

All of the code used to produce every analysis presented in this paper
and the manuscript itself is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences and assembled transcriptomes used are deposited in SRA, and
new sequences for this work can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in
each species with ortholog annotations are located in the supplemental
data file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM}
values are in all\_species\_per\_sample\_fpkm.txt.xz.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \ac{ANXA1}, \ac{ANXA2},
\ac{S100A10}, \ac{S100A11}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A4, S100A6, S100A10, and S100A11
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\ac{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads}.

@@

The evolutionarily convergent capture of endogenous retroviruses to
form expressed fusogenic proteins such as Syncytin-1 and -2
(\ac{ERVW-1} and \ac{ERVFRD-1}, respectively) has been implicated in
syncytiotrophoblast formation in \human and many other
species~\cite{Rawn.ea2008:evolutionregulationandfunction}. While genes
which are similar to \ac{ERVFRD-1} and \ac{ERVW-1} are expressed in
most of the placentas analyzed here, there are no highly expressed
genes with sequence similarity in \bonobo and \dog, even though these
mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
and the syncytiotrophoblast must be continuously replenished, the
gene(s) responsible for cytotrophoblast fusion to form the
syncytiotrophoblast should be highly expressed in the core placenta.
\ac{ANXA5} is known to be expressed at multiple stages of
placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. We hypothesize that \ac{ANXA5} is essential
for the completion of fusion and the maintenance of membrane repair
during the flux of nutrients and waste in the placenta
\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing,Ueki.ea2012:Lossofmaternalannexin}.

@@

This study includes only term placentas, and our conclusions are
limited to the set of genes which continue to be expressed at term. We
suspect that many genes which are responsible for establishing shape,
intimacy, and interdigitation are expressed and function at earlier
time points where placental morphology is being established.
Furthermore, while these samples have been dissected to only include
the fetal side, they likely include substantial cell and tissue
heterogeneity. We are therfore not yet able to assign specific genes
in our core placental transcriptome to specific cell types and
tissues. We believe that samples of placentas from earlier in
gestation and from multiple locations within the placenta and at
multiple will address these limitations.

@@

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance and genes responsible
for cell-cell and cell-matrix interactions, such as the annexins and \ac{LGALS1}. These
genes were almost certainly present in the placenta of the most recent
common mammalian ancestor of the species included in this study. We hypothesize that
modifications to the expression or function of genes in the core placenta transcriptome
will compromise the
development and organization of the placenta.
